---
title: "Merging full count 1940 census from IPUMS with CenSoc Data"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Merging IPUMS and Censoc

It is straightforward to match CENSOC data to IPUMS census data once both datasets have been [obtained](https://censoc.demog.berkeley.edu/articles/ipums_document.html).  

The variable "HISTID" uniquely identifies all records for both Censoc and the full count IPUMS census datasets and can be used to match records between the two datasets. 

A tidyverse join: 

```{r, eval = F}
library(tidyverse)
censoc <- read_csv('path/to/your/censoc')
census <- read_csv('path/to/your/ipums')
merged_df <- censoc %>%
  left_join(census, by = "HISTID")
```

A data.table merge (probably faster): 
```{r, eval = F}
library(data.table)
censoc <- fread('path/to/your/censoc')
census <- fread('path/to/your/ipums')
setkey(censoc, HISTID)
setkey(census, HISTID)
merged_df <- censoc[census, nomatch=0]
```

Using the [IPUMSR Package] (https://cran.r-project.org/web/packages/ipumsr/vignettes/ipums.html): 
```{r, eval = F}
library(ipumsr)
library(tidyverse)

read_ipums_micro(data_file = "path/to/your/ipums",  ddi = "path/to/your/ipums/ddi")
censoc <- read_csv('path/to/your/censoc')

merged_df <- censoc %>%
  left_join(census, by = "HISTID")
```
