---
title: "Merging IPUMS and Censoc"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Merge

It is straightforward to match CENSOC data to IPUMS census data once both datasets have been [obtained](https://censoc.demog.berkeley.edu/articles/ipums_document.html). Instructions in this Vignette pertain to the V2 Version of the CenSoc File. 

The variable "HISTID" uniquely identifies all records for both CenSoc and the full count IPUMS census datasets and can be used to match records between the two datasets. 

Base R (no packages required):

```{r, eval = FALSE}
censoc <- read.csv('path/to/your/censoc')
census <- read.csv('path/to/your/ipums')
merged_df <- merge(censoc, census, by = "HISTID")
```


Merge using "tidyverse" packages (faster): 

```{r, eval = FALSE}
library(tidyverse)
censoc <- read_csv('path/to/your/censoc')
census <- read_csv('path/to/your/ipums')
merged_df <- censoc %>%
  inner_join(census, by = "HISTID")
```

Merge using data.table (fastest): 

```{r, eval = FALSE}
library(data.table)
censoc <- fread('path/to/your/censoc')
census <- fread('path/to/your/ipums')
setkey(censoc, HISTID)
setkey(census, HISTID)
merged_df <- censoc[census, nomatch=0]
```

The full IPUMS 1940 Census might be too large for some computer's memory. The following code uses the [IPUMSR Package](https://cran.r-project.org/web/packages/ipumsr/vignettes/ipums.html) to read in the census file in "chunks", match each "chunk" the censoc file, and combine the results of each chunk (this approach includes value labels for IPUMS variables!): 

```{r, eval = FALSE}
library(ipumsr)
library(tidyverse)

censoc <- read_csv("path/to/your/censoc")

cb_function <- function(x, pos) {
  x %>% 
    inner_join(censoc, by = "HISTID")
}

cb <- IpumsDataFrameCallback$new(cb_function)

## This will take a while. For info on downloading IPUMS ddi, please see: 
# https://cran.r-project.org/web/packages/ipumsr/vignettes/ipums.html
merged_df <- read_ipums_micro_chunked(
  ddi = "/path/to/your/census/ddi/", data_file = "/path/to/your/census/data/", verbose = FALSE,
  callback = cb, chunk_size = 5000000
)
```


## Characteristics of Merged Dataset

The merged dataset should contain 7,563,270 records. For IPUMS Full Count 1940 extracts made after January 31st, 2019, an additional 39 records in the CenSoc V2 can't be matched onto the IPUMS 1940 Full Count File. These 39 records are still included in the CenSoc V2 file.

Checking that you can replicate these descriptive statisics with your merged dataset is a helpful way to ensure the merge was successful.

#### Race: 

```{r, eval = F}
merged_df %>%
                 group_by(RACE) %>%
                 tally() %>%
                 mutate(freq = signif(n*100 / sum(n), 3)) %>%
                 select(RACE, n, freq)
```


| RACE|       n|    freq|
|----:|-------:|-------:|
|    1| 7104150| 93.9000|
|    2|  435239|  5.7500|
|    3|   12268|  0.1620|
|    4|    3873|  0.0512|
|    5|    5566|  0.0736|
|    6|    2174|  0.0287|

#### Average age at death by race: 

```{r, eval = F}
# code the race variable to have meaningful names, see: https://usa.ipums.org/usa-action/variables/RACE#codes_section
merged_df <- merged_df %>% mutate(race_name = case_when(
  RACE == 1 ~ "White",
  RACE == 2 ~ "Black",
  RACE == 3 ~ "American Indian/Alaskan native",
  RACE == 4 ~ "Chinese",
  RACE == 5 ~ "Japanese",
  RACE == 6 ~ "Other Asian or Pacific Islander",
  TRUE ~ "NA"
))

# calculate age at death
merged_df <- merged_df %>%
  mutate(age_at_death = dyear - byear)

# calculate average age at death by race
merged_df %>% group_by(race_name) %>% 
  summarise(mean_age_at_death = mean(age_at_death))
```

|race_name                       | mean_age_at_death|
|:-------------------------------|-----------------:|
|American Indian/Alaskan native  |          71.39501|
|Black                           |          73.48693|
|Chinese                         |          77.36328|
|Japanese                        |          76.93388|
|Other Asian or Pacific Islander |          78.94296|
|White                           |          76.08481|

